Nitric-oxide-feedback to ciliary photoreceptor cells orchestrates UV avoidance

### 2023-01-01 text/Manuscript.Rmd
merge

NO generally acts via soluble guanylate cyclases (sGC), belonging to the guanylate cyclase family with a CYC domain (PF00211). NO binding to the heme group of sGC leads to increased cyclic guanosine monophosphate (cGMP) production. Analysis of sGCs in *Platynereis* indicated that these genes are not expressed in any of the cells of the cPRC circuit ([@Veraszto2017] and data not shown). Recently, Moroz and coworkers reported an atypical but widely conserved family of guanylyl cyclases with a NIT (nitrite/nitrate sensing) domain (PF08376) (NIT-GC) as potential mediators of NO signalling [@Moroz2020]. To search for NIT-GCs in _Platynereis_, we searched transcriptome resources and retrieved 15 potential NIT-GC homologs. To analyse the relationship of these sequences to metazoan NIT-GCs, we retrieved protein sequences with a CYC domain from the transcriptome and genome databases of 45 metazoan species and [**how many?] choanoflagellate. We carried out cluster analysis and did phylogenetic reconstruction on a group of membrane-bound guanylyl cyclases with sGCs as an outgroup. In agreement with Moroz et al. [@Moroz2020],we found a group of GCs with NIT domains with representatives in placozoans, cnidarians, some ecdysozoans, echinoderms, and lophotrochozoans. The 15 _Platynereis_ sequences were parts of several major and ancient clades of the tree (Figure 4---figure supplement 1). 
To characterise the expression of NIT-GCs, we using previously published spatially mapped single-cell transcriptome data [@Achim2015; @williams2017]. Among the 15 NIT-GCs, four showed high and specific expression in the cPRCs and one was expressed in the INNOS cells (Figure 3---figure supplement 1). In the single-cell data, we could identify the cPRCs by the specific expression of c-opsin1 and a pedal-peptide2 neuropeptide precursor (MLD proneuropeptide) [@Arendt2004; @williams2017], which have previously been described as cPRC markers (Figure 3A and B, Figure 3---figure supplement 2). The INNOS cells were identified by NOS expression and their spatial mapping. 
We decided to focus on two NIT-GCs expressed in the cPRCs and with a full-length sequences, NIT-GC1 and NIT-GC2. To confirm the single-cell data, we first carried out HCR with probes for _NIT-GC1_ and _NIT-GC2_ mRNA. Both genes showed very clear and specific expression in the four cPRCs, as confirmed by co-labeling with an acetylated tubulin antibody and with an HCR probe against _MLD proneuropeptide_. To analyse the subcellular localisation of NIT-GC1 and NIT-GC2 at the protein level, we raised and affinity-purified polyclonal antibodies against a specific peptide sequence from both proteins. In immunostainings, we found that NIT-GC1 was localise to the region corresponding to the axonal projections of the cPRCs in the anterior nerve plexus. Co-immunostaining with the rabbit NIT-GC1 and the rat NOS antibodies revealed that both proteins were localised in close proximity in the neurosecretory plexus (Figure 3---figure supplement 3). In contrast, NIT-GC2 strongly and specifically labelled the large sensory ciliary region of the cPRCs (Figure 3C and D). These different subcellular localisations suggest that the two NIT-GCs are involved in different intracellular signalling processes in the ciliary and axonal regions of the cPRCs.

>>>>

NO generally acts via soluble guanylate cyclases (sGC), belonging to the guanylate cyclase family with a CYC domain (PF00211). NO binding to the heme group of sGC leads to increased cyclic guanosine monophosphate (cGMP) production. Analysis of sGCs in *Platynereis* indicated that these genes are not expressed in any of the cells of the cPRC circuit ([@Veraszto2017] and data not shown). Recently, Moroz and coworkers reported an atypical but widely conserved family of guanylyl cyclases with a NIT (nitrite/nitrate sensing) domain (PF08376) (NIT-GC) as potential mediators of NO signalling [@Moroz2020]. To search for NIT-GCs in *Platynereis*, we searched transcriptome resources and retrieved 15 potential NIT-GC homologs. To analyse the relationship of these sequences to metazoan NIT-GCs, we retrieved protein sequences with a CYC domain from the transcriptome and genome databases of 45 metazoan species and [\*\*how many?] choanoflagellate. We carried out cluster analysis and did phylogenetic reconstruction on a group of membrane-bound guanylyl cyclases with sGCs as an outgroup. In agreement with Moroz et al. [@Moroz2020],we found a group of GCs with NIT domains with representatives in placozoans, cnidarians, some ecdysozoans, echinoderms, and lophotrochozoans. The 15 *Platynereis* sequences were parts of several major and ancient clades of the tree (Figure 4---figure supplement 1). To characterise the expression of NIT-GCs, we using previously published spatially mapped single-cell transcriptome data [@Achim2015; @williams2017]. Among the 15 NIT-GCs, four showed high and specific expression in the cPRCs and one was expressed in the INNOS cells (Figure 3---figure supplement 1). In the single-cell data, we could identify the cPRCs by the specific expression of c-opsin1 and a pedal-peptide2 neuropeptide precursor (MLD proneuropeptide) [@Arendt2004; @williams2017], which have previously been described as cPRC markers (Figure 3A and B, Figure 3---figure supplement 2). The INNOS cells were identified by NOS expression and their spatial mapping. We decided to focus on two NIT-GCs expressed in the cPRCs and with a full-length sequences, NIT-GC1 and NIT-GC2. To confirm the single-cell data, we first carried out HCR with probes for *NIT-GC1* and *NIT-GC2* mRNA. Both genes showed very clear and specific expression in the four cPRCs, as confirmed by co-labeling with an acetylated tubulin antibody and with an HCR probe against *MLD proneuropeptide*. To analyse the subcellular localisation of NIT-GC1 and NIT-GC2 at the protein level, we raised and affinity-purified polyclonal antibodies against a specific peptide sequence from both proteins. In immunostainings, we found that NIT-GC1 was localise to the region corresponding to the axonal projections of the cPRCs in the anterior nerve plexus. Co-immunostaining with the rabbit NIT-GC1 and the rat NOS antibodies revealed that both proteins were localised in close proximity in the neurosecretory plexus (Figure 3---figure supplement 3). In contrast, NIT-GC2 strongly and specifically labelled the large sensory ciliary region of the cPRCs (Figure 3C and D). These different subcellular localisations suggest that the two NIT-GCs are involved in different intracellular signalling processes in the ciliary and axonal regions of the cPRCs.
